Shaping strategy in core and flake industry：Standardized manufacturing of QSY (North China) faceted spheroids

### 2025-05-20 analysis/paper/paper.qmd
edits to methods

In this study, the 3d models of the cores were collected and used for morphometric and technological analysis, which not only improved the accuracy of the data but also facilitated scholars to verify and replicate the results. Digital data was captured using Artec spider, and post-processing was carried out with Artec studio.

@@

Based on typological study, we identify 13 rounded faceted sub-spheroids/spheroids in QSY site. Inevitably, the classification of stone artifacts under the guidance of typology relies on subjective judgment, and the identification of stone core types may vary with researchers.

To avoid preconceptions and self-justification, we try to apply quantitative method to help confirm the spheroids in site. These morphological measurements are designed to indicate whether these identified spheroids are spherical and the differences or similarities between them and standard spheres. We will also examine the degree of standardization of the identified spheroids.

In addition to identified spheroids, other multidirectional-multifacial cores will also be included in the morphological study to explore whether they can be distinguishable from each other. Furthermore, we will compare the results obtained by typology and morphometric analysis. The specific methods with details are introduced below and the code is also available:

@@

To evaluate the degree to which the artifacts approximate a perfect sphere, this study employs a standardized sphericity index originally proposed by [@Wadell1935] and later adapted for 3D digital models in geoscientific research. The index is calculated using the following formula:

@@

Where V is the volume and A is the surface area of the object. This dimensionless metric yields a maximum value of 1 for a perfect sphere. Values closer to 1 indicate a higher degree of sphericity, thus serving as a reliable proxy for assessing the morphological regularity and intentional shaping of the spheroids. All surface area and volume measurements were derived from 3D mesh models that were decimated to a consistent face count and subjected to unit normalization, ensuring comparability across specimens.

@@

Spherical Harmonic (SPHARM) analysis is a method for describing 3D shapes using a series of mathematical functions defined on the surface of a sphere. It works like a 3D version of Fourier analysis, breaking down complex shapes into components of different spatial frequencies. Each shape is first mapped onto a unit sphere and then expressed as a combination of spherical harmonic functions $Y_l^m(\theta, \phi)$. The shape is represented by a set of coefficients $\hat{f}(l, m)$, which capture how much each function contributes. Here, θ and ϕ are angles in spherical coordinates, l and m are order and degree respectively. Lower-order terms describe the general shape, while higher-order terms capture finer details[@Link2024; @Medyukhina2020; @Harper2022].

This method is widely used in the field of biology, such as for the analysis of skull morphology and cell morphology, but there are currently few attempts in archaeological artifacts [@Muller2023; @Noshita2025]. However, as we mentioned when introducing the method, SPHARM is especially useful for shape analysis of objects that are roughly spherical in shape, such as the spheroids and cores studied in this research.

All analyses are conducted in Python with fully reproducible code. First, STL files of spheroids and cores are imported, cleaned, and simplified to extract 3D mesh vertices, which are converted from Cartesian (x, y, z) to spherical coordinates (r, θ, φ). Next, radial values are interpolated onto a regular spherical grid and then decomposed using spherical harmonics based on the Driscoll–Healy sampling scheme. The resulting coefficients are normalized for scale-invariant comparison across models.

From the normalized spherical harmonic coefficients, we compute the rotation-invariant power spectrum by summing the squared magnitudes of the coefficients at each degree. Each 3D shape is thus represented by its unique power spectrum, which serves as a multidimensional descriptor. To explore variation between models, we apply UMAP for dimensionality reduction and visualization [@Van2023].

This study represents the first attempt in archaeology to apply SPHARM for the classification of cores. Specifically, we include a set of standard geometric models—such as sphere, rounded cube, ellipsoid (1.2:1:0.8) rounded rectangular prism (1.2:1:1), and disc-like (flattened, and like discoid core)—in the SPHARM analysis and UMAP. Cores are positioned in this position based on their morphological similarity to these reference models: individuals that cluster more closely to a specific standard form exhibit corresponding shape characteristics.

>>>>

In this study, the 3d models of the cores were collected and used for morphometric and technological analysis, ~~which not only improved the accuracy of the data but also facilitated scholars to verify and replicate the results~~. Digital data were captured using Artec Spider, and post-processing was carried out with Artec Studio [version number]. The 3D scan files and all the R and Python code used in the analyses reported here are openly available online at https://doi.org/xxxx/xxxx/.

@@

Based on typological study, we identify 13 rounded faceted sub-spheroids/spheroids in QSY site. Inevitably, the classification of stone artifacts under the guidance of typology relies on subjective judgment, and the identification of stone core types may vary with researchers. [BM: feels like this belongs in results section]

~~To avoid preconceptions and self-justification,~~ We ~~try to~~ applied quantitative methods to ~~help confirm the spheroids in site. These morphological measurements designed to~~ indicate ~~whether these identified spheroids are spherical and~~ the degree of difference between the archaeological spheroids and a perfect sphere. This included computing the sphericity index, 3D spherical harmonics, spherical harmonic energy and the centroid offset for each artefact. We ~~will~~ also examined the degree of standardization of the spheroids by comparing coefficients of variation [citation] of sphericity [and what else].

In addition to the archaeological spheroids, we included other multidirectional-multifacial cores in the morphological study to explore whether they can be distinguished from the spheroids. Furthermore, we compared the results of the typological and our new morphometric analyses to investigate the validity of the traditional typological approach.

@@

To evaluate the degree to which the artifacts approximate a perfect sphere, this study employs a standardized sphericity index originally proposed by @Wadell1935 and later adapted for 3D digital models in geoscientific research. The index is calculated using the following formula:

@@

Where V is the volume and A is the surface area of the object. This dimensionless metric yields a maximum value of 1 for a perfect sphere. Values closer to 1 indicate a higher degree of sphericity, thus serving as a reliable proxy for assessing the morphological regularity and intentional shaping of the spheroids. We used Python to derive surface area and volume measurements from 3D mesh models. The models were were decimated to a consistent face count and subjected to unit normalization, ensuring comparability across specimens. [BM: check if we need this here]

@@

Spherical Harmonic (SPHARM) analysis is a method for describing 3D shapes using a series of mathematical functions defined on the surface of a sphere. It works like a 3D version of Fourier analysis, breaking down complex shapes into components of different spatial frequencies. Each shape is first mapped onto a unit sphere and then expressed as a combination of spherical harmonic functions $Y_l^m(\theta, \phi)$. The shape is represented by a set of coefficients $\hat{f}(l, m)$, which capture how much each function contributes. Here, θ and ϕ are angles in spherical coordinates, l and m are order and degree respectively. Lower-order terms describe the general shape, while higher-order terms capture finer details[@Link2024; @Medyukhina2020; @Harper2022]. This method is widely used in the field of biology, such as ~~for~~ the analysis of skull morphology and cell morphology [citations]. However, there are currently very few application of SPHARM to archaeological artifacts [@Muller2023; @Noshita2025]. Our work aims to demonstrate the usefulness of SPHARM for shape analysis of objects that are roughly spherical in shape, such as the spheroids and cores studied in this research.

We first imported, cleaned, and simplified the STL files of spheroids and cores to extract 3D mesh vertices, which were then converted from Cartesian (x, y, z) to spherical coordinates (r, θ, φ) [citation]. Next, we interpolated radial values onto a regular spherical grid and then decomposed them using spherical harmonics based on the Driscoll–Healy sampling scheme [citation]. The resulting coefficients were normalized for scale-invariant comparison across models.

From the normalized spherical harmonic coefficients, we computed the rotation-invariant power spectrum by summing the squared magnitudes of the coefficients at each degree. Each 3D shape is thus represented by its unique power spectrum, which serves as a multidimensional descriptor. To explore variation between models, we applied UMAP for dimensionality reduction and visualization [@Van2023].

This study represents the first attempt in archaeology to apply SPHARM for the classification of cores. Specifically, we include a set of standard geometric models—such as sphere, rounded cube, ellipsoid (1.2:1:0.8) rounded rectangular prism (1.2:1:1), and disc-like (flattened, and like discoid core)—in the SPHARM analysis and UMAP. Cores are positioned in this position based on their morphological similarity to these reference models: individuals that cluster more closely to a specific standard form exhibit corresponding shape characteristics. [BM: this text probably better in the intro or background section]
